Formal Grammars as Types in Non-commutative Bunched Type Theory

### 2023-12-27 paper/paper.tex
some notes on canonicity

\documentclass[acmsmall,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}

\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\Set}{\mathbf{Set}}

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
\newcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}

@@

% TODO: add Pedro's info

@@

\begin{abstract}
  We propose the category of functions from strings to sets as a
  universal syntax-independent model of formal grammars, generalizing
  the set of strings model of formal language theory to account for
  ambiguity. The morphisms of this category give a notion of parse
  transformer, and isomorphism corresponds to \emph{strong
  equivalence} of grammars. We observe that the standard algebraic
  constructions of regular expressions and their extension to
  mu-regular expressions correspond to universal constructions in this
  category, which we axiomatize as a structure we call a 2-Kleene
  Algebra.

\end{abstract}

@@

\section{Future work}

>>>>

\documentclass[sigconf,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{stmaryrd}

\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}

@@

@@

\begin{abstract}
  We propose the category of functions from strings to sets as a
  universal syntax-independent model of formal grammars, generalizing
  the set of strings model of formal language theory to account for
  ambiguity. The morphisms of this category give a notion of parse
  transformer, and isomorphism corresponds to \emph{strong
  equivalence} of grammars. We observe that the standard algebraic
  constructions of regular expressions and their extension to
  mu-regular expressions correspond to universal constructions in this
  category, which we axiomatize as a structure we call a 2-Kleene
  Algebra.
\end{abstract}

\maketitle

@@

\section{Universal Constructions in the Category of Grammars}

Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
and therefore certain initial algebras

\section{A Bunched Type Theory for Grammars}

\subsection{Semantics}

Set types $A$ and set contexts $\Delta$ denotes sets, and
grammar types $\gamma$ and grammar contexts $\Gamma$ in context
$\Delta$ denote families of grammars $(\Set^{\Sigma^*})^\Delta$.  A

\subsection{Canonicity}

Let $\phi$ be a propositional BI formula over $\Sigma$. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $\phi$:
\begin{enumerate}
\item On the one hand we have the semantics just defined, $\llbracket \phi \rrbracket w$
\item On the other hand, we can view the string $w$ as a formula
  $\lceil w \rceil$ and define a parse to be an equivalence classes of
  proofs $x:\lceil w \rceil \vdash \phi$
\end{enumerate}

It is not difficult to see that at least for the ``strictly positive''
formulae (those featuring only the positive connectives
$\vee,\bot,\epsilon,\otimes,\mu$ and the generators $c$) that every
element $t \in \llbracket \phi \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the language. However it is far less clear that every proof
$x:\lceil w \rceil \vdash p : \phi$ corresponds to some sequence of
introduction forms since proofs can include elimination forms as
well. So we want to prove a kind of \emph{canonicity} result for the
calculus, a generalization of the disjunction theorem of constructive
logic:

\begin{enumerate}
\item Every proof $\lceil w \rceil \vdash p : \phi \otimes \psi$ is equal to $(q,r)$ for some $q,r$
\item Every proof $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
\item Every proof $\lceil w \rceil \vdash p : \phi \vee \psi$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
\item Every proof $\lceil w \rceil \vdash p : c$ is equal to $x:c$
\item Every proof $\lceil w \rceil \vdash p : \mu X. \phi$ is equal to $\textrm{roll}(q)$ where $q : \phi(\mu X.\phi/X)$
\end{enumerate}

To prove this result we will use a simple Kripke logical relation.

First, let $L$ be the category of BI formulae and proofs (quotiented
by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
\[ N(\phi)(w) = L(w,\phi) \]

Then define the gluing category $\mathcal G$ as the comma category
$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
a formula $\phi \in L$ and an object $S \in \mathcal
P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
description of such an $S$: it is simply a family of sets indexed by
proofs $L(w,\phi)$:
\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
This category clearly comes with a projection functor $\pi : \mathcal
G \to \mathcal L$ and then our goal is to define a section by using
the universal property of $\mathcal L$.

To this end we define
\begin{enumerate}
\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
\item $\mu X. ??$ ??
\end{enumerate}

\section{Logical Characterizations of Language Classes}

Our bunched type theory

Predicate BI with existential quantifiers is already enough to define
formulae that correspond to arbitrary recursively enumerable languages
since we can simulate the tape of a Turing machine using two
stacks. I.e., we can define a set $S$ of lists of tape symbols, and
then define a formula that first ``copies'' the input onto one stack, then executes

\[ \mu \textrm{Copy} : S \to *. \lambda s. (\epsilon \wedge T\,s\,[])\bigvee_{c \in \Sigma} c (S\,{c::s}) \]

\section{Related Work}

TODO

%% That paper with Fritz Henglein about regexes as types

%% Vaughan Pratt on the residual operator
\section{Future work}
